|

### 2021-12-15 01-Chp-Background.Rmd
Ciara edits

The chapter introduces evidence synthesis as the key framework used to guide the research presented in the remaining chapters. Finally, it outlines the aims, objectives and structure of this thesis, and briefly summarises the contributions to the scientific literature that arose from this research.

@@

One of the most commonly used sets of criteria for diagnosing dementia cases are the Diagnostic and Statistical Manual of Mental Disorders (DSM) criteria (Table \@ref(tab:diagnosticCriteria-table)).[@edition2013] Outlined in Table \@ref(tab:diagnosticCriteria-table), these form the broad definition of a dementia diagnoses, and are supported by a detailed patient history, evidence from carers and family members, and objective assessments of cognitive ability using neurocognitive tests.

Many cognitive assessment tools exist for the purpose of informing a diagnoses of dementia,[@sheehan2012] with two of the best known of these being the Mini Mental State Exam (MMSE) and Montreal Cognitive Assessment (MoCA) scale. The distinction between these memory scales and diagnostic criteria presented above should be noted. For example, the MMSE is used to provide evidence for part A of the criteria presented in \@ref(tab:diagnosticCriteria-table). Taken alone, it does not indicate the absence or presence of dementia, instead merely indicating cognitive impairment which could be due to another cause (for example, temporary delirium as a result of an infection or surgery).

@@

<!----------------------------------------------------------------------------->
(ref:diagnosticCriteria-caption) __Overview of the DSM-5 criteria for dementia and vascular dementia.__ - These published criteria[@edition2013] are commonly used to defined dementia outcomes. <!-- TODO Need to finish this sentence and explan criteria better -->

(ref:diagnosticCriteria-scaption) Overview of the DSM-5 criteria for dementia and vascular dementia.

@@

Differentiating between the underlying causes of a dementia diagnosis is challenging but necessary, as whether the patient has Alzheimer's disease or vascular dementia will affect expected progression and potential treatment options available (see Section \@ref(intro-treatments)). Cause-specific criteria exist for the diagnosis of dementia sub-types. For example, the NINCDS-ADRDA criteria are commonly used to assess patients for Alzheimer's disease,[@dubois2007] while vascular dementia is diagnosed using the NINCDS-AIREN criteria.[@roman1993]

@@

Dementia is quickly becoming a critically important public health issue. Despite the age-specific incidence and prevalence of dementia remaining relatively constant over time,[@prince2016] an ageing population is set to create a dementia epidemic, particularly in Westernised countries.[@flier2005] While approximately 525,000 patients have received a dementia diagnosis, the true number of people currently living with dementia in the UK is thought to be closer to 850,000, with this figure expected to double by 2040.[@baker2019] Globally, the prevalence of dementia is expected to reach 75 million by 2030.[@prince2016] Dementia is a leading cause of death in the UK,[@zotero-15757] and one of the few without a proven treatment.

@@

Triglycerides are the simplest and most common type of lipids found across the body. They are used to store unused calories from food, and to move energy around the body.[@laufs2020] In contrast, cholesterol is primarily used to create cell walls and certain sex hormones.[@zampelas2019] As lipids are not water soluble, within the blood stream, cholesterol is transported in lipoprotein structures of varying densities. Low-density-lipoprotein cholesterol (LDL-c), commonly know as the "bad" cholesterol, transports fat to cells, acting as an energy conveyor. In contrast, High density-lipoprotein cholesterol (HDL-c), transports cholesterol to the liver to be broken down and excreted.[@feingold2000]

@@

Several statin treatments have been widely available for some time (see Table \@ref(tab:statinOverview-table)). Depending on the statin and dosage prescribed, the average reduction in LDL-c concentrations ranges from 15% with low-intensity regimen (e.g. ravastatin 5 mg/day) up to 60% with a high-intensity regimen (e.g. rosuvastatin 80 mg/day).[@collins2016; @law2003] Statins also vary with regard to their lipophilicity (the extent to which they are lipid soluble), affecting their localisation within the body, with hydophilic statins being concentrated in the liver and lipophilic statins circulating more widely.[@schachter2005] This may create a divide in the pleiotropic affects of statins with differing lipophilicity, particularly given the ability of lipophilic statins to permeate the blood brain barrier.[@sierra2011]

@@

A role for lipids in the aetiology of dementia is supported by both genetic linkage studies and functional cell biology studies. The generation of the amyloid plaques found in the brains of Alzheimer's patients is cholesterol dependent,,[@burns2003; @mizuno1999] while the most established genetic risk factor for late-onset dementia, apolipoprotein E (_Apo_$\mathcal{E}$), is involved in cerebral cholesterol transport. Several other genes involved in cholesterol transport have also been found to be associated with increased AD susceptibility.[@beecham2014; @harold2009; @meng2007]

@@

By far the largest source of evidence on the relationship between blood lipids and dementia outcomes comes from observational designs. Several studies have examined the relationships between concentrations of serum lipids (total cholesterol (TC), low density lipoprotein cholesterol (LDL-c), high density lipoprotein cholesterol (HDL-c) and triglycerides) <!-- TODO These abbreviations have already been defined above --> and both Alzheimer's disease and vascular dementia and reported extremely varied results. In some studies, a high serum cholesterol concentration has been found to be associated with an increase in susceptibility to AD,[@kivipelto2002; @kivipelto2005; @schilling2017; @solomon2009; @whitmer2005] however others have shown no association,[@li2005; @mainous2005; @mielke2010; @tan2003] or a reduced susceptibility [@mielke2005; @reitz2004]. With regard to vascular dementia, decreased levels of HDL-c appear to be associated with increased risk [@reitz2004], while for LDL-c, studies have reported both positive and negative associations.[@reitz2004; @moroney1999]

@@

In terms of the central research of this thesis, RCTs of statin therapy can be used to provide indirect evidence for the effect of reducing blood LDL-c levels on dementia risk. However, RCTs may be infeasible if the outcome of interest is one with a long prodromal period, such as dementia (see Section \@ref(underlying-pathologies)), as they would require extremely long and costly follow-up.[@ritchie2015] It is no surprise then that the two previous trials providing evidence on the effect of statins on dementia risk, identified by a recent Cochrane review,[@mcguinness2016] are in fact trials of statins for the prevention of coronary related outcomes.

While being widely cited, these studies have major limitations that reduce their utility as a source of evidence on the effect of statin treatment in assessing the impact of lipid-lowering treatment on dementia risk. Firstly, the criteria used to determine a dementia outcome are unclear. One of the trials, the Prospective Study of Pravastatin in the Elderly (PROSPER) trial,[@trompet2010] reported not on dementia outcomes but on the change in cognitive scores over a mean of 3.2 years. As highlighted in Section \@ref(diagnostic-criteria), a "change in score" alone is insufficient to diagnose a dementia outcome. The second trial, the Medical Research Council/British Health Foundation Protection Study,[@heartprotectionstudycollaborativegroup2002] found no effect of simvastatin on dementia (`r estimate(1, 0.61, 1.65)`), but did not report how the outcome was assessed/recorded within the trial.

Additionally, the two trials did not make any effort to assign an underlying pathology to each case, instead reporting an all-cause dementia outcome. As discussed in Section \@ref(underlying-pathologies), the different underlying pathologies of dementia have different mechanisms of action, and so it is not gauranteed that the effect of statins would be consistent across them.

Both trials were also limited by the relatively short follow-up period examined. This limited follow-up was expected, as the primary aim of both trials was to assess the effect of statins on short-term coronary-related outcomes.[@heartprotectionstudycollaborativegroup2002; @trompet2010] The PROSPER trial had a mean follow-up of 3.2 years, while the MRC/BHF Protection Study estimated risk at 5 years of follow-up. Given the long lag time between non-symptomatic onset of dementia and clinical presentation, it is likely that these durations are insufficient to fully capture the onset of dementia. Finally, as they included only patients at high vascular risk, their generalisability to other settings is limited.[@mcguinness2016]

@@

Newer methodological approaches, such as Mendelian randomisation (MR),[@daveysmith2014] have also been used to examine the effect of varying lipid levels on dementia risk. Mendelian randomisation attempts to combat the risk of reverse causation and residual confounding inherent to observational studies by using natural random varation in participants genomes.[@greenland2000] In brief, MR uses genetic variants that are both strongly associated with the exposure of interest and are independent from potential confounders to strengthen causal inference.[@daveysmith2014] The analytic method relies on several assumptions about the instrumental variable (IV),[@davies2018] namely that:

@@

Recent MR studies indicated that genetically determined low levels of LDL-c may cause a reduction in AD risk.[@larsson2017; @ostergaard2015] However, the effect was attenuated in sensitivity analysis that exclude the region surrounding the _Apo_$\mathcal{E}$ gene, the strongest known risk factor for Alzheimer's disease.[@kim2009] Inclusion of _Apo_$\mathcal{E}4$ variants invalidates the exclusion restriction criteria (Assumption 3, above), as the risk reduction observed may be driven by variants in this region via a pathway independent of lipid levels. This was supported by further MR studies where _Apo_$\mathcal{E}4$ variants were intentionally excluded.[@benn2017] Despite the increasing number of MR studies examining this topic, no systematic review of this study design as a source of evidence has been performed.

@@

Evidence synthesis is the process of finding and integrating information from several sources to examine a research question.[@donnelly2018] A common type of evidence synthesis is a systematic review, either with or without a meta-analysis.[@chandler2019chapter]

The results of an evidence synthesis exercise can be used to provide a more definitive answer to that question or, failing that, to highlight gaps in the existing evidence base. The ability to identify these gaps is particularly useful in guiding future research to address questions that have yet to be answered.

This thesis seeks to use an evidence synthesis framework to assess the effect of lipids, and treatments that influence lipid levels, on dementia outcomes. Specifically, this thesis considers three concepts within the umbrella term of evidence synthesis:

@@

However, the role of preprints in evidence syntheses is less well defined. Defined by the Committee on Publication Ethics as 'scholarly manuscript[s] posted by the author(s) in an openly accessible platform, usually before or in parallel with the peer review process',[@committeeonpublicationethicscope2018] preprints serve several purposes. They are used to establish primacy when submitting to a journal where the peer-review process may take several months.[@vale2016] In addition, the allow for the rapid disseminate research findings, as occurred during the COVID-19 pandemic.[@fraser2020preprinting] Finally, they are a source of information on manuscripts that may not have been accepted elsewhere, helping to combat publication bias or the "file-drawer" effect.[@rosenthal1979]

@@

In this thesis, preprints are considered an important source of evidence, in contrast to previous reviews on this topic. However, as with grey literature,[@mahood2014] there are several logistical problems with carrying out systematic searches in preprint repositories. As such, to enable the inclusion of preprints in the systematic review described in Chapters \@ref(sys-rev-methods-heading), a new tool addressing these issues is presented in Chapter \@ref(sys-rev-tools-heading).

@@

In the context of this thesis, if lipids are found to have a causal role in development of dementia, evidence-based preventative strategies would be best informed by identifying the types of individuals who are most likely to receive benefit from treatment with lipid-modifying agents.[@arain2009; @clayton2018; @mccartney2016] However, if primary studies do not present results stratified by covariates of interest, meta-analyses of summary-level data on this topic often have limited ability to examine research questions related to exposure-covariate interactions.[@riley2010] In terms of this thesis, participant age and sex are considered to be of particular interest.[@arain2009; @letenneur1999]

An IPD meta-analysis of lipid levels on dementia outcomes would overcome this limitation of summary-level data, as access to the raw data allows for an analysis that investigates these interactions.[@riley2020] This approach has the added benefit of allowing a common set of inclusion criteria and a common statistical model to be applied across all datasets, potentially eliminating some important sources of heterogeneity.[@stewart2002]

Despite their advantages, IPD meta-analysis are rarely performed.[@tugwell2010] Factors limiting their uptake include the increased time and effort they require when compared to a summary-level analysis, and the low success rate associated with obtaining the raw data.[@nevitt2017; @ventresca2020] The data underlying primary studies are frequently not publicly available,[@alsheikh-ali2011; @federer2018] and the availability of data "available on request from authors" declines rapidly over time.[@vines2014] Several systematic barriers to open data sharing have been identified[@vanpanhuis2014]. Of particular concern for biomedical IPD analyses are legal issues surrounding the sharing of medical data, motivated by concerns around patient privacy.[@wartenberg2010]

@@

I will attempt to obtain the raw data from relevant primary studies identified by the systematic review in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading). Any data obtained will be combined with that available from the DPUK portal as part of an IPD meta-analysis in Chapter \@ref(ipd-heading). This will enable the assessment of the effect of lipids on dementia stratified by key variables such as sex.

<!----------------------------------------------------------------------------->
### Triangulating across study designs {#intro-triangulation}

As illustrated in Section \@ref(evidence-association), several diverse epidemiological methods have been used to examine the effect of varying blood lipid levels on dementia risk. However, each method is limited by its own biases. Aetiological triangulation is a developing evidence synthesis method that seeks to exploit these inherent differences in study design, and as a result, in biases.[@lawlor2016] If several sources of evidence are available and point towards identical conclusions about an exposure-outcome relationship, and these sources are at risk of unrelated biases, this strengthens our confidence in the result. The ideal scenario is where predicted sources of bias are likely to be in competing directions, one to strengthen the effect of the exposure and the other to attenuate it.[@lawlor2016] As such, triangulating these results can provide a middle-ground between the competing directions of bias. A triangulation approach can also prove useful in helping to prospectively design new studies that are at risk of different sources of bias to those already observed in the published literature.[@munafo2018] However, existing methods for triangulation are limited to a qualitative discussion of the different identified results with respect to their biases,[@lawlor2016] an approach which faces issues in interpretability when assessing many results.

To address this limitation, this thesis seeks to develop and apply a novel quantitative triangulation approach to answer causal questions on the effect of blood lipids on dementia outcomes. All existing evidence, regardless of study design, is first identified and assessed using domain-based tools by the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading). The existing evidence base is then supplemented by two primary analyses, presented in Chapters \@ref(cprd-analysis-heading) and \@ref(ipd-heading). Finally all existing and new evidence is incorporated into the quantiatative triangulation framework in Chapter \@ref(tri-heading).

@@

### Aims and objectives

The over-arching aim of this thesis is to explore the relationship between blood lipid levels, and by extension treatments that modify blood lipid levels such as statins, and the subsequent risk of dementia and related outcomes

The specific research objectives that this thesis seeks to address are:

@@

Chapters are self-contained, presenting the methods and results of that specific research project. They are bookended by introductory and discussion sections which place the methods and results in context. Each chapter is prefaced by a "lay" or plain English summary. <!-- TODO developed with input from the Patient and Public Advisory Group (see Section \@ref(disc-PPI) for a discussion of the group's involvement and Appendix \@ref(appendix-ppi) for more detail on the group). -->

-   **Chapter \@ref(background-heading):** Background information on dementia and blood lipid levels. This chapter provides an introduction to the topics covered in this thesis to non-subject area experts, and discusses the motivation for the remainder of the thesis.
-   **Chapter \@ref(sys-rev-tools-heading):** This chapter introduces a new tool, `medrxivr`, which was used to developed to allow for systematic searches of the health-related preprint repositories.
-   **Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading):** These chapters describe, respectively, the methods and results of a comprehensive systematic review and meta-analysis. This review examined all available evidence on the effect of blood lipids and interventions that modified blood lipids on dementia outcomes.
-   **Chapter \@ref(cprd-analysis-heading):** This chapter examines the relationship between lipid-regulating agent use and dementia outcomes in the Clinical Practice Research Datalink, a large primary care electronic health record database.
-   **Chapter \@ref(ipd-heading):** This chapter describes an individual participant data analysis of several longitudinal cohort studies to describe the relationship between blood serum lipids and dementia outcomes, stratified by important covariates such as sex.
-   **Chapter \@ref(tri-heading)**: This chapter integrates the evidence identified and produced by the preceeding chapters as part of a quantitative triangulation framework.
-   **Chapter \@ref(discussion-heading)**: This chapter summarise the key clinical and methodological findings of the thesis and discusses their implications for practice. The overall strengths and weaknesses of this project are discussed in detail, and further avenues of research are suggested.

@@

### Contributions to the scientific literature {#literature-contributions}

@@

The experience of extracting data for the systematic review in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading) inspired a practical guide for researchers. This piece was co-written with Dr. Emily Hennessy (see Author Declarations in the front materials).

@@

The tool used to visualise the risk-of-bias assessments in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading) has been published in Research Synthesis Methods. See Appendix \@ref(appendix-robvis) for more details on this tool. Note that this publication does not describe the recently-developed functionality for producing bias direction plots, as described in Chapter \@ref(tri-heading).

@@

An R package containing function to implement the quantitative triangulation approach detailed in Chapter \@ref(tri-heading). At present, only the development version of the package is available, which can be installed from GitHub using:

@@

* In this introductory chapter, I provided background information on the core elements of the central research question and framed the research presented in the context of a theoretical framework of evidence synthesis. I outlined the central aim of the thesis and provided an overview of the individual analyses performed.

* I also described the contributions of this thesis to the scientific literature. Finally, I discussed the research tools and software developed to support the analyses in this thesis, one of which (the preprint search tool) is introduced more fully in the following chapter.

>>>>

It then introduces evidence synthesis as the key framework used to guide the research presented in the remaining chapters. Finally, it outlines the aims, objectives and structure of this thesis, and briefly summarises the contributions to the scientific literature that arose from this research.

@@

One of the most commonly used sets of criteria for diagnosing dementia cases are the Diagnostic and Statistical Manual of Mental Disorders (DSM) criteria (Table \@ref(tab:diagnosticCriteria-table)).[@edition2013] Outlined in Table \@ref(tab:diagnosticCriteria-table), these form the broad definition of a dementia diagnosis, and are supported by a detailed patient history, evidence from carers and family members, and objective assessments of cognitive ability using neurocognitive tests.

Many cognitive assessment tools exist for the purpose of informing a diagnoses of dementia,[@sheehan2012] with two of the best known of these being the Mini Mental State Exam (MMSE) and Montreal Cognitive Assessment (MoCA) scale. The distinction between these memory scales and diagnostic criteria presented above should be noted. For example, the MMSE is used to provide evidence for part A of the criteria presented in \@ref(tab:diagnosticCriteria-table). Taken alone, it does not indicate the absence or presence of dementia, instead merely indicating cognitive impairment which could have another cause (for example, temporary delirium as a result of an infection or surgery).

@@

<!----------------------------------------------------------------------------->
(ref:diagnosticCriteria-caption) __Overview of the DSM-5 criteria for dementia (major neurocognitive disorder).__ - These published criteria[@edition2013] are commonly used to define dementia cases. Criterion A captures the cognitive decline aspect of dementia, while Criteria B-D ensure that the impairment observed is sufficient to interfere with daily living and can not explained by other causes. Note that all criteria (A-D) must be met.

(ref:diagnosticCriteria-scaption) Overview of the DSM-5 criteria for dementia (major neurocognitive disorder)

@@

Differentiating between the underlying causes of a dementia diagnosis is challenging but necessary, because whether the patient has Alzheimer's disease or vascular dementia will affect expected progression and potential treatment options available (see Section \@ref(intro-treatments)). Cause-specific criteria exist for the diagnosis of dementia sub-types. For example, the NINCDS-ADRDA criteria are commonly used to assess patients for Alzheimer's disease,[@dubois2007] while vascular dementia is diagnosed using the NINCDS-AIREN criteria.[@roman1993]

@@

Dementia is quickly becoming a critically important public health issue. Despite the age-specific incidence and prevalence of dementia remaining relatively constant over time,[@prince2016] an ageing population is set to create a dementia epidemic, particularly in westernised countries.[@flier2005] While approximately 525,000 patients have received a dementia diagnosis, the true number of people currently living with dementia in the UK is thought to be closer to 850,000, with this figure expected to double by 2040.[@baker2019] Globally, the prevalence of dementia is expected to reach 75 million by 2030.[@prince2016] Dementia is a leading cause of death in the UK,[@zotero-15757] and one of the few without a proven treatment.

@@

Triglycerides are the simplest and most common type of lipids, used to store excess calories from food and to move energy around the body.[@laufs2020] In contrast, cholesterol is primarily used to create cell walls and certain sex hormones.[@zampelas2019] Lipids are not water soluble, and so cholesterol is transported through the blood stream in lipoprotein structures of varying densities. Low-density-lipoprotein cholesterol (LDL-c), commonly know as "bad" cholesterol, acts as an energy conveyor by transporting fat to cells. In contrast, high density-lipoprotein cholesterol (HDL-c) transports cholesterol to the liver to be broken down and excreted.[@feingold2000]

@@

Several statin treatments have been widely available for some time (see Table \@ref(tab:statinOverview-table)). Depending on the statin and dosage prescribed, the average reduction in LDL-c concentrations ranges from 15% with low-intensity regimen (e.g. ravastatin 5 mg/day) up to 60% with a high-intensity regimen (e.g. rosuvastatin 80 mg/day).[@collins2016; @law2003] Statins also vary with regard to their lipophilicity (the extent to which they are lipid soluble), affecting their localisation within the body, with hydophilic statins being concentrated in the liver and lipophilic statins circulating more widely.[@schachter2005] This may create a divide in the off-target effects of statins with differing lipophilicity, particularly given the ability of lipophilic statins to permeate the blood brain barrier.[@sierra2011]

@@

A role for lipids in the aetiology of dementia is supported by both genetic linkage studies and functional cell biology studies. The generation of the amyloid plaques found in the brains of Alzheimer's patients is cholesterol dependent,[@burns2003; @mizuno1999] while the most established genetic risk factor for late-onset dementia, apolipoprotein E (_Apo_$\mathcal{E}$), is involved in cerebral cholesterol transport. Several other genes involved in cholesterol transport have also been found to be associated with increased AD susceptibility.[@beecham2014; @harold2009; @meng2007]

@@

By far the largest source of evidence on the relationship between blood lipids and dementia outcomes comes from observational designs. Several studies have examined the relationships between concentrations of serum lipids (total cholesterol (TC), LDL-c, HDL-c and triglycerides) and both Alzheimer's disease and vascular dementia. These studies have reported extremely varied results. In some, a high serum cholesterol concentration has been found to be associated with an increase in susceptibility to AD,[@kivipelto2002; @kivipelto2005; @schilling2017; @solomon2009; @whitmer2005] however others have shown no association,[@li2005; @mainous2005; @mielke2010; @tan2003] or a reduced susceptibility [@mielke2005; @reitz2004]. With regard to vascular dementia, decreased levels of HDL-c appear to be associated with increased risk [@reitz2004], while for LDL-c, studies have reported both positive and negative associations.[@reitz2004; @moroney1999]

@@

In terms of the central research of this thesis, randomised controlled trials (RCTs) of statin therapy can be used to provide indirect evidence on the effect of reducing blood LDL-c levels on dementia risk. However, RCTs may be infeasible if the outcome of interest is one with a long prodromal period, such as dementia (see Section \@ref(underlying-pathologies)), as they would require extremely long and costly follow-up.[@ritchie2015] It is no surprise then that the two previous trials providing evidence on the effect of statins on dementia risk, identified by a recent Cochrane review,[@mcguinness2016] are in fact trials of statins for the prevention of coronary-related outcomes.

While being widely cited, these studies have major limitations that reduce their utility as a source of evidence on the effect of statin treatment in assessing the impact of lipid-lowering treatment on dementia risk. Firstly, the criteria used to determine a dementia outcome are prone to misclassification. One of the trials, the Prospective Study of Pravastatin in the Elderly (PROSPER) trial,[@trompet2010] reported not on dementia outcomes but on the change in cognitive scores. As alluded to in Section \@ref(diagnostic-criteria), a "change in score" alone is insufficient to diagnose a dementia outcome. The second trial, the Medical Research Council/British Health Foundation Protection Study,[@heartprotectionstudycollaborativegroup2002] found no effect of simvastatin on dementia (`r estimate(1, 0.61, 1.65)`), but did not report how the outcome was assessed/recorded within the trial.

Additionally, the two trials did not make any effort to assign an underlying pathology to each case, instead reporting an all-cause dementia outcome. As discussed in Section \@ref(underlying-pathologies), the different underlying pathologies of dementia have different mechanisms of action, and so it is not guaranteed that the effect of statins would be consistent across them.

Both trials were also limited by the relatively short follow-up period examined. This limited follow-up was to be expected because the primary aim of both trials was to assess the effect of statins on short-term coronary-related outcomes.[@heartprotectionstudycollaborativegroup2002; @trompet2010] The PROSPER trial had a mean follow-up of 3.2 years, while the MRC/BHF Protection Study estimated risk at 5 years of follow-up. Given the long lag time between non-symptomatic onset of dementia and clinical presentation, it is likely that these durations are insufficient to fully capture the onset of dementia. Finally, as they included only patients at high vascular risk, their generalisability to other settings is limited.[@mcguinness2016]

@@

Newer methodological approaches, such as Mendelian randomisation (MR),[@daveysmith2014] have also been used to examine the effect of varying lipid levels on dementia risk. Mendelian randomisation attempts to combat the risk of reverse causation and residual confounding inherent to observational studies by using natural random varation in participants' genomes.[@greenland2000] In brief, MR uses genetic variants that are both strongly associated with the exposure of interest and are independent from potential confounders to strengthen causal inference.[@daveysmith2014] The analytic method relies on several assumptions about the instrumental variable (IV),[@davies2018] namely that:

@@

Recent MR studies indicated that genetically determined low levels of LDL-c may cause a reduction in AD risk.[@larsson2017; @ostergaard2015] However, the effect was attenuated in sensitivity analyses that exclude the region surrounding the _Apo_$\mathcal{E}$ gene, the strongest known risk factor for Alzheimer's disease.[@kim2009] Inclusion of _Apo_$\mathcal{E}$ variants invalidates the exclusion restriction criteria (Assumption 3, above), as the risk reduction observed may be driven by variants in this region via a pathway independent of lipid levels. This was supported by further MR studies in which the observed effect attenuated to the null after _Apo_$\mathcal{E}$ variants were excluded.[@benn2017] Despite the increasing number of MR studies examining this topic, no systematic review of this study design as a source of evidence has been performed.

@@

Evidence synthesis is the process of finding and integrating information from several sources to examine a research question.[@donnelly2018] The results of an evidence synthesis exercise can be used to provide a more definitive answer to that question or, failing that, to highlight gaps in the existing evidence base. The ability to identify these gaps is particularly useful in guiding future research to address questions that have yet to be answered. A common type of evidence synthesis exercise is a systematic review, either with or without a meta-analysis.[@chandler2019chapter]

This thesis seeks to use an evidence synthesis framework to assess the effect of lipid levels, and treatments that influence lipid levels, on dementia outcomes. Specifically, this thesis considers three concepts within the umbrella term of evidence synthesis:

@@

However, the role of preprints in evidence synthesis is less well defined. Described by the Committee on Publication Ethics as 'scholarly manuscript[s] posted by the author(s) in an openly accessible platform, usually before or in parallel with the peer review process',[@committeeonpublicationethicscope2018] preprints serve several purposes. They are used to establish primacy when submitting to a journal where the peer-review process may take several months.[@vale2016] In addition, the allow for the rapid dissemination of research findings, as occurred during the COVID-19 pandemic.[@fraser2020preprinting] Finally, they are a source of information on manuscripts that may not have been accepted elsewhere, helping to combat publication bias or the "file-drawer" effect.[@rosenthal1979]

@@

In this thesis, preprints are considered an important source of evidence, in contrast to previous reviews on this topic. However, as with grey literature,[@mahood2014] there are several logistical problems with carrying out systematic searches in preprint repositories. As such, to enable the inclusion of preprints in the systematic review described in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading), a new tool addressing these issues is presented in Chapter \@ref(sys-rev-tools-heading).

@@

In the context of this thesis, if lipids are found to have a causal role in development of dementia, evidence-based preventative strategies would be best informed by identifying the types of individuals who are most likely to receive benefit from treatment with lipid-modifying agents.[@arain2009; @clayton2018; @mccartney2016] However, if primary studies do not present results stratified by covariates of interest, meta-analyses of summary-level data on this topic often have limited ability to examine research questions related to exposure-covariate interactions.[@riley2010] Based on existing evidence, participant age and sex are considered to be of particular interest.[@arain2009; @letenneur1999]

An IPD meta-analysis of lipid levels on dementia outcomes would overcome this limitation of summary-level data because access to the raw data allows for an analysis that investigates these interactions.[@riley2020] This approach has the added benefit of allowing a common set of inclusion criteria and a common statistical model to be applied across all datasets, potentially eliminating some important causes of heterogeneity.[@stewart2002]

Despite their advantages, IPD meta-analyses are rarely performed.[@tugwell2010] Factors limiting their uptake include the increased time and effort they require when compared to a summary-level analysis, and the low success rate associated with obtaining the raw data.[@nevitt2017; @ventresca2020] The data underlying primary studies are frequently not publicly available,[@alsheikh-ali2011; @federer2018] and the availability of data "available on request" from authors declines rapidly over time.[@vines2014] Several systematic barriers to open data sharing have been identified[@vanpanhuis2014]. Of particular concern for biomedical IPD analyses are legal issues surrounding the sharing of medical data, motivated by concerns around patient privacy.[@wartenberg2010]

@@

I will attempt to obtain the raw data from relevant primary studies identified by the systematic review presentd in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading). Any data obtained will be combined with that available from the DPUK portal as part of an IPD meta-analysis in Chapter \@ref(ipd-heading). This will enable the assessment of the effect of lipids on dementia, and whether this relationship is modified by key variables such as participants' age or sex.

### Triangulating across study designs {#intro-triangulation}

As illustrated in Section \@ref(evidence-association), several diverse epidemiological methods have been used to examine the effect of varying blood lipid levels on dementia risk. Aetiological triangulation is a developing evidence synthesis method that seeks to exploit these differences in study design, and as a result, in underlying bias structures.[@lawlor2016] If several sources of evidence are available and point towards identical conclusions about an exposure-outcome relationship, and these sources are at risk of unrelated biases, this strengthens our confidence in the result. The ideal scenario is where predicted sources of bias are likely to be in competing directions, one to strengthen the effect of the exposure and the other to attenuate it.[@lawlor2016] As such, triangulating these results can provide a middle-ground between the competing directions of bias. A triangulation approach can also prove useful in helping to prospectively design new studies that are at risk of different sources of bias to those already observed in the published literature.[@munafo2018] However, existing methods for triangulation are limited to a qualitative discussion of the different identified results with respect to their biases,[@lawlor2016] an approach which faces issues in interpretability when assessing many results.

To address this limitation, this thesis seeks to develop and apply a novel quantitative triangulation approach to answer causal questions on the effect of blood lipids on dementia outcomes. All existing evidence, regardless of study design, is first identified and assessed for risk of bias by the systematic review presented in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading). The existing evidence base is then supplemented by two primary analyses, presented in Chapters \@ref(cprd-analysis-heading) and \@ref(ipd-heading). Finally all existing and new evidence is incorporated into the quantitative triangulation framework in Chapter \@ref(tri-heading).

@@

### Aim and objectives

The central aim of this thesis was to infer the causal effect of blood lipid levels on dementia outcomes via evidence synthesis methods. The specific research objectives that this thesis seeks to address are:

@@

Chapters are self-contained, presenting the methods and results of that specific research project. The sole exception to this is the systematic review, which due to its size is split across Chapters \@ref(sys-rev-methods-heading) (Methods) & \@ref(sys-rev-results-heading) (Results). Each chapter is bookended by introductory and summary sections which place the methods and results in context. Finally, each chapter is prefaced by a "lay" or plain English summary.

-   **Chapter \@ref(background-heading):** Background information on dementia and blood lipid levels. This chapter provides an introduction to the topics covered in this thesis, and discusses the motivation for the research presented in the following chapters.
-   **Chapter \@ref(sys-rev-tools-heading):** This chapter introduces a new tool, `medrxivr`, which was developed to allow for systematic searches of the health-related preprint repositories.
-   **Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading):** These chapters describe, respectively, the methods and results of a comprehensive systematic review and meta-analysis. This review examined all available evidence on the association of blood lipids and LRA with subsequent risk of dementia.
-   **Chapter \@ref(cprd-analysis-heading):** This chapter examines the relationship between lipid-regulating agent use and dementia outcomes in the Clinical Practice Research Datalink, a large primary care electronic health record database.
-   **Chapter \@ref(ipd-heading):** This chapter describes an individual participant data analysis of three longitudinal cohort studies to describe the relationship between blood lipids and dementia outcomes
-   **Chapter \@ref(tri-heading)**: This chapter integrates the evidence identified and produced by the preceding chapters as part of a quantitative triangulation framework.
-   **Chapter \@ref(discussion-heading)**: This chapter summarise the key clinical and methodological findings of the thesis and discusses their implications for practice. The overall strengths and weaknesses of this project are discussed in detail, and further avenues of research are suggested.

@@

### Contributions to the literature {#literature-contributions}

@@

The experience of extracting data for the systematic review in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading) inspired a practical guide which aims to help primary researchers make their data synthesis-ready. This piece was co-written with Dr. Emily Hennessy (see Acknowledgements in the front materials).

@@

A publication describing the tool used to visualise the risk-of-bias assessments in Chapters \@ref(sys-rev-methods-heading) & \@ref(sys-rev-results-heading) has been published in Research Synthesis Methods. See Appendix \@ref(appendix-robvis) for more details on this tool. Note that this publication does not describe the recently-developed functionality for producing bias direction plots, as described in Chapter \@ref(tri-heading).

@@

An R package containing functionality to implement the quantitative triangulation approach detailed in Chapter \@ref(tri-heading). At present, only the development version of the package is available, which can be installed from GitHub using:

@@

* In this introductory chapter, I provided background information on the core elements of the central research question and framed the research presented in the context of a theoretical framework of evidence synthesis.

* I also outlined the central aim of this thesis and provided an overview of the individual analyses performed. I described the contributions of this thesis to the field, both in terms of peer-review publications and conference abstracts.

* Finally, I discussed the research tools and software developed to support the analyses performed in this thesis, one of which (the preprint search tool) is introduced more fully in the following chapter.
